Continous time Markov chain models; \\ extension, simulation and visualisation

### 2015-03-06 appendices/appendixc.tex
fixing floats
implementing comments
tiny table
changes for printout

\section{Introduction} % LM: ? lacking a better name

@@

%%
\begin{figure}[H]
\centering
\includegraphics[scale=\figurescale]{LOAD_TREE.png} 
\caption{
{ \footnotesize 
{\bf How to load a tree into {\bussname}.}
} % END: footnotesize
}
\label{fig:loadingtopology}
\end{figure}
%%

@@

%%
\begin{figure}[H]
\centering
\includegraphics[scale=\figurescale]{DEMOGRAPHIC.png} 
\caption{
{ \footnotesize 
{\bf Choosing a demographic (coalescent) model.}
} % END: footnotesize
}
\label{fig:demographic}
\end{figure}
%%

@@

%%
\begin{figure}[H]
\centering
\includegraphics[scale=\figurescale]{HKY.png} 
\caption{
{ \footnotesize 
{\bf Setting a HKY substitution model with $\kappa$ parameter of $1.0$.}
} % END: footnotesize
}
\label{fig:hky}
\end{figure}
%%

@@

%%
\begin{figure}[H]
\centering
\includegraphics[scale=\figurescale]{GY94.png} 
\caption{
{ \footnotesize 
{\bf Setting a codon model (GY94) with $\kappa=1.0$ and $\omega=0.1$.}
} % END: footnotesize
}
\label{fig:gy94}
\end{figure}
%%

@@

%%
\begin{figure}[H]
\centering
\includegraphics[scale=\figurescale]{GAMMA.png} 
\caption{
{ \footnotesize 
{\bf Site rate heterogeinity.}
Site heterogeinity can be simulated by going to the \textbf{Partitions} and click on the `Site Rate Model' button.
} % END: footnotesize
}
\label{fig:gamma}
\end{figure}
%%

@@

%%
\begin{figure}[H]
\centering
\includegraphics[scale=\figurescale]{BASE_FREQUENCIES.png} 
\caption{
{ \footnotesize 
{\bf Base frequencies.}
To specify base frequencies, go to the \textbf{Partitions} table and click the `Frequency Model' button.
} % END: footnotesize
}
\label{fig:bases}
\end{figure}
%%

@@

%%
\begin{figure}[H]
\centering
\includegraphics[scale=\figurescale]{CLOCK2.png} 
\caption{
{ \footnotesize 
{\bf Clock model.}
Setting up an Inverse Gaussian relaxed clock with mean$=0$, standard deviation of $1.0$ and no offset.
} % END: footnotesize
}
\label{fig:clock2}
\end{figure}
%%

@@

%%
\begin{figure}[H]
\centering
\includegraphics[scale=\figurescale]{MULTIPLE_PARTITIONS.png} 
\caption{
{ \footnotesize 
{\bf Setting multiple partitions.}
You can set up any number of partitions, each with its own tree topology, model of evolution, molecular clock and base frequency. 
Just click the `+' button at the bottom left corner of the GUI window at the \textbf{Partitions} tab.
} % END: footnotesize
}
\label{fig:partitions}
\end{figure}
%%

@@

% FB: how to color cells in multirow table?
% TODO: LM: Update the table to show AA options
\begin{table}[H]
\begin{center}
\footnotesize{
\begin{tabular}{lcc}
\hline 
\textbf{Parameter} & \textbf{Command} & \textbf{Options}\tabularnewline
\hline 
\cellcolor{snow3}Tree topology & \cellcolor{snow3}-treeFile & \cellcolor{snow3}-\tabularnewline
\hline 
Taxa set & -taxaSet & -\tabularnewline
\hline 
\cellcolor{snow3} & \cellcolor{snow3} & \cellcolor{snow3}-constantPopulationParameterValues\tabularnewline
\cellcolor{snow3} & \cellcolor{snow3} & \cellcolor{snow3}-exponentialGrowthRateParameterValues\tabularnewline
\cellcolor{snow3}\multirow{-3}{*}{Demographic (coalescent) model} & \cellcolor{snow3}\multirow{-3}{*}{-demographicModel} & \cellcolor{snow3}-exponentialDoublingTimeParameterValues\tabularnewline
\hline 
 &  & -HKYsubstitutionParameterValues\tabularnewline
 &  & -GTRsubstitutionParameterValues\tabularnewline
 &  & -TN93substitutionParameterValues\tabularnewline
\multirow{-4}{*}{Markov model of evolution} & \multirow{-4}{*}{-branchSubstitutionModel} & -GY94substitutionParameterValues\tabularnewline
\hline 
\cellcolor{snow3}Site rate heterogeinity & \cellcolor{snow3}-siteRateModel & \cellcolor{snow3}-gammaSiteRateModelParameterValues\tabularnewline
\hline 
 &  & -strictClockParameterValues\tabularnewline
 &  & -lognormalRelaxedClockParameterValues\tabularnewline
 &  & -exponentialRelaxedClockParameterValues\tabularnewline
\multirow{-4}{*}{Molecular clock model} & \multirow{-4}{*}{-clockRateModel} & -inverseGaussianRelaxedClockParameterValues\tabularnewline
\hline 
\cellcolor{snow3} & \cellcolor{snow3} & \cellcolor{snow3}-nucleotideFrequencyParameterValues\tabularnewline
\cellcolor{snow3}\multirow{-2}{*}{Base frequencies} & \cellcolor{snow3}\multirow{-2}{*}{-baseFrequencies} & \cellcolor{snow3}-codonFrequencyParameterValues\tabularnewline
\hline 
Root sequence & -rootSequence & -\tabularnewline
\end{tabular}
} % END: footnotesize
\caption{
{ \footnotesize 
{\bf List of command line options for {\bussname}.}
} % END: footnotesize
}
\label{tab:commands}
\end{center}
\end{table}

@@

%%
\begin{figure}[H]
\centering
\includegraphics[scale=\figurescale]{SIMUTAB.png} 
\caption{
{ \footnotesize 
{\bf Setting RNG seed.}
To set the seed for your simulation and choose the parallel BEAGLE implementation, check the corresponding tick boxes at the {\bf Simulation} tab.
} % END: footnotesize
}
\label{fig:simutab}
\end{figure}
%%

@@

\begin{figure}[H]
\centering
\includegraphics[scale=\figurescale]{CODON_FREQUENCIES.png} 
\caption{
{ \footnotesize 
{\bf Setting codon frequencies.}
Just select codon frequencies in the drop-down menu and you're done. The default is $1/61$ for each codon.
} % END: footnotesize
}
\label{fig:CodonFrequencies}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[scale=\figurescale]{CODON_PARTITIONING.png} 
\caption{
{ \footnotesize 
{\bf Codon partitioned simulation.}
} % END: footnotesize
}
\label{fig:CodonPartitoning}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[scale=\figurescale]{TAXA_EDITOR.png} 
\caption{
{ \footnotesize 
{\bf Loading a taxa set from tab-separated file.}
Once you load the taxa set onto {\bussname}, you are ready to simulate under any of the coalescent dynamics available.
} % END: footnotesize
}
\label{fig:TaxaEditor}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[scale=\figurescale]{DEMOGRAPHIC_MODEL.png} 
\caption{
{ \footnotesize 
{\bf Loading a taxa set from tab-separated file.}
} % END: footnotesize
}
\label{fig:DemoModel}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[scale=\figurescale]{TERMINAL_OUTPUT.png} 
\caption{
{ \footnotesize 
{\bf Terminal output.}
} % END: footnotesize
}
\label{fig:Terminal}
\end{figure}

@@

We need to define the discrete geographical locations that {\bussname} will sample from. After the  {\color{darkblue}taxa} block paste the following code:

@@

Scroll to the {\color{darkblue}frequencyModel} block. Because we have specified $K=3$ discrete locations we need to specify base frequencies for them. Delete the existing {\color{darkblue}frequencyModel} block and paste there following code instead:

@@

Now scroll further down to the substitution model block (defined between {\color{darkblue}HKYModel} tags if you generated the XML using default settings). 
We need to define a geographical substitution model, driven by a rate matrix. 
We specify $K\times(K-1)=6$ entries such that the underlying instantaneous rate matrix is asymmetric (non-reversible). 
Replace the existing block with:

@@

Now reference the created substitution model inside {\color{darkblue}siteModel} tags like this:

@@

Finally all the new elements need to be referenced in the {\color{darkblue}beagleSequenceSimulator} block. We specify one replicate, as every taxa can ultimately have only one geographical trait:

@@

@@

\subsection{License}
% LM: licence info I found at BEAST source 
This is free software; you can redistribute it and/or modify it under the terms of the {GNU} Lesser General Public License as published by the Free Software Foundation; either version 2 of the License, or (at your option) any later version. This software is distributed in the hope that it will be useful, but {WITHOUT ANY WARRANTY}; without even the implied warranty of {MERCHANTABILITY} or {FITNESS FOR A PARTICULAR PURPOSE}. See the ``GNU Lesser General Public License'':\url{http://www.gnu.org/licenses/lgpl.html} for more details.

>>>>

\section{Introduction}

@@

%%
\begin{figure}[h!]
\centering
\includegraphics[scale=\figurescale]{LOAD_TREE.png} 
\caption{
{ \footnotesize 
{\bf How to load a tree into {\bussname}.}
} % END: footnotesize
}
\label{fig:loadingtopology}
\end{figure}
%%

@@

%%
\begin{figure}[h!]
\centering
\includegraphics[scale=\figurescale]{DEMOGRAPHIC.png} 
\caption{
{ \footnotesize 
{\bf Choosing a demographic (coalescent) model.}
} % END: footnotesize
}
\label{fig:demographic}
\end{figure}
%%

@@

%%
\begin{figure}[h!]
\centering
\includegraphics[scale=\figurescale]{HKY.png} 
\caption{
{ \footnotesize 
{\bf Setting a HKY substitution model with $\kappa$ parameter of $1.0$.}
} % END: footnotesize
}
\label{fig:hky}
\end{figure}
%%

@@

%%
\begin{figure}[h!]
\centering
\includegraphics[scale=\figurescale]{GY94.png} 
\caption{
{ \footnotesize 
{\bf Setting a codon model (GY94) with $\kappa=1.0$ and $\omega=0.1$.}
} % END: footnotesize
}
\label{fig:gy94}
\end{figure}
%%

@@

%%
\begin{figure}[h!]
\centering
\includegraphics[scale=\figurescale]{GAMMA.png} 
\caption{
{ \footnotesize 
{\bf Site rate heterogeinity.}
Site heterogeinity can be simulated by going to the \textbf{Partitions} and click on the `Site Rate Model' button.
} % END: footnotesize
}
\label{fig:gamma}
\end{figure}
%%

@@

%%
\begin{figure}[h!]
\centering
\includegraphics[scale=\figurescale]{BASE_FREQUENCIES.png} 
\caption{
{ \footnotesize 
{\bf Base frequencies.}
To specify base frequencies, go to the \textbf{Partitions} table and click the `Frequency Model' button.
} % END: footnotesize
}
\label{fig:bases}
\end{figure}
%%

@@

%%
\begin{figure}[h!]
\centering
\includegraphics[scale=\figurescale]{CLOCK2.png} 
\caption{
{ \footnotesize 
{\bf Clock model.}
Setting up an Inverse Gaussian relaxed clock with mean$=0$, standard deviation of $1.0$ and no offset.
} % END: footnotesize
}
\label{fig:clock2}
\end{figure}
%%

@@

%%
\begin{figure}[h!]
\centering
\includegraphics[scale=\figurescale]{MULTIPLE_PARTITIONS.png} 
\caption{
{ \footnotesize 
{\bf Setting multiple partitions.}
You can set up any number of partitions, each with its own tree topology, model of evolution, molecular clock and base frequency. 
Just click the `+' button at the bottom left corner of the GUI window at the \textbf{Partitions} tab.
} % END: footnotesize
}
\label{fig:partitions}
\end{figure}
%%

@@

\begin{table}[h!]
\begin{center}
\tiny{
\begin{tabular}{lcc}
\hline 
\textbf{Parameter} & \textbf{Command} & \textbf{Options}\tabularnewline
\hline 
\cellcolor{snow3}Tree topology & \cellcolor{snow3}-treeFile & \cellcolor{snow3}-\tabularnewline
\hline 
Taxa set & -taxaSet & -\tabularnewline
\hline 
\cellcolor{snow3} & \cellcolor{snow3} & \cellcolor{snow3}-constantPopulationParameterValues\tabularnewline
\cellcolor{snow3} & \cellcolor{snow3} & \cellcolor{snow3}-exponentialGrowthRateParameterValues\tabularnewline
\cellcolor{snow3}\multirow{-3}{*}{Demographic (coalescent) model} & \cellcolor{snow3}\multirow{-3}{*}{-demographicModel} & \cellcolor{snow3}-exponentialDoublingTimeParameterValues\tabularnewline
\hline 
 &  & -HKYsubstitutionParameterValues\tabularnewline
 &  & -GTRsubstitutionParameterValues\tabularnewline
 &  & -TN93substitutionParameterValues\tabularnewline
\multirow{-4}{*}{Markov model of evolution} & \multirow{-4}{*}{-branchSubstitutionModel} & -GY94substitutionParameterValues\tabularnewline
\hline 
\cellcolor{snow3}Site rate heterogeinity & \cellcolor{snow3}-siteRateModel & \cellcolor{snow3}-gammaSiteRateModelParameterValues\tabularnewline
\hline 
 &  & -strictClockParameterValues\tabularnewline
 &  & -lognormalRelaxedClockParameterValues\tabularnewline
 &  & -exponentialRelaxedClockParameterValues\tabularnewline
\multirow{-4}{*}{Molecular clock model} & \multirow{-4}{*}{-clockRateModel} & -inverseGaussianRelaxedClockParameterValues\tabularnewline
\hline 
\cellcolor{snow3} & \cellcolor{snow3} & \cellcolor{snow3}-nucleotideFrequencyParameterValues\tabularnewline
\cellcolor{snow3}\multirow{-2}{*}{Base frequencies} & \cellcolor{snow3}\multirow{-2}{*}{-baseFrequencies} & \cellcolor{snow3}-codonFrequencyParameterValues\tabularnewline
\hline 
Root sequence & -rootSequence & -\tabularnewline
\end{tabular}
} % END: footnotesize
\caption{
{ \footnotesize 
{\bf List of command line options for {\bussname}.}
} % END: footnotesize
}
\label{tab:commands}
\end{center}
\end{table}

@@

%%
\begin{figure}[h!]
\centering
\includegraphics[scale=\figurescale]{SIMUTAB.png} 
\caption{
{ \footnotesize 
{\bf Setting RNG seed.}
To set the seed for your simulation and choose the parallel BEAGLE implementation, check the corresponding tick boxes at the {\bf Simulation} tab.
} % END: footnotesize
}
\label{fig:simutab}
\end{figure}
%%

@@

\begin{figure}[h!]
\centering
\includegraphics[scale=\figurescale]{CODON_FREQUENCIES.png} 
\caption{
{ \footnotesize 
{\bf Setting codon frequencies.}
Just select codon frequencies in the drop-down menu and you're done. The default is $1/61$ for each codon.
} % END: footnotesize
}
\label{fig:CodonFrequencies}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[scale=\figurescale]{CODON_PARTITIONING.png} 
\caption{
{ \footnotesize 
{\bf Codon partitioned simulation.}
} % END: footnotesize
}
\label{fig:CodonPartitoning}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[scale=\figurescale]{TAXA_EDITOR.png} 
\caption{
{ \footnotesize 
{\bf Loading a taxa set from tab-separated file.}
Once you load the taxa set onto {\bussname}, you are ready to simulate under any of the coalescent dynamics available.
} % END: footnotesize
}
\label{fig:TaxaEditor}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[scale=\figurescale]{DEMOGRAPHIC_MODEL.png} 
\caption{
{ \footnotesize 
{\bf Loading a taxa set from tab-separated file.}
} % END: footnotesize
}
\label{fig:DemoModel}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[scale=\figurescale]{TERMINAL_OUTPUT.png} 
\caption{
{ \footnotesize 
{\bf Terminal output.}
} % END: footnotesize
}
\label{fig:Terminal}
\end{figure}

@@

We need to define the discrete geographical locations that {\bussname} will sample from. After the  {\color{darkblue}\mbox{taxa}} block paste the following code:

\medskip{}

@@

\medskip{}

Scroll to the {\color{darkblue}\mbox{frequencyModel}} block. Because we have specified $K=3$ discrete locations we need to specify base frequencies for them. Delete the existing {\color{darkblue}\mbox{frequencyModel}} block and paste there following code instead:

\medskip{}

@@

\medskip{}

Now scroll further down to the substitution model block (defined between {\color{darkblue}\mbox{HKYModel}} tags if you generated the XML using default settings). 
We need to define a geographical substitution model, driven by a rate matrix. 
We specify $K\times(K-1)=6$ entries such that the underlying instantaneous rate matrix is asymmetric (non-reversible). 
Replace the existing block with:

\medskip{}

@@

\medskip{}

Now reference the created substitution model inside {\color{darkblue}\mbox{siteModel}} tags like this:

\medskip{}

@@

\medskip{}

Finally all the new elements need to be referenced in the {\color{darkblue}\mbox{beagleSequenceSimulator}} block. We specify one replicate, as every taxa can ultimately have only one geographical trait:

\medskip{}

@@

\medskip{}

@@

\subsection{License}
% LM: licence info I found at BEAST source 
This is free software; you can redistribute it and/or modify it under the terms of the {GNU} Lesser General Public License as published by the Free Software Foundation; either version 2 of the License, or (at your option) any later version. This software is distributed in the hope that it will be useful, but {WITHOUT ANY WARRANTY}; without even the implied warranty of {MERCHANTABILITY} or {FITNESS FOR A PARTICULAR PURPOSE}. See the ``GNU Lesser General Public License'' (\url{http://www.gnu.org/licenses/lgpl.html}) for more details.

### 2015-03-06 chapters/chapter04.tex
implementing comments
changes for printout

%%%%%%%%%%%%%%5

\noindent
\textbf{Availability and requirements:} \\
The parallel BEAST/BEAGLE utility for sequence simulation is licensed under the GNU Lesser GPL and its source code is freely available as part of the BEAST Google Code repository: \url{www.code.google.com/p/beast-mcmc/}
Compiled, runnable packages targeting all major platforms along with a tutorial and supplementary data are hosted at:
\url{www.rega.kuleuven.be/cev/ecv/software/pibuss}.
{\bussname} requires Java runtime environment version 1.5 or later to run its executables and the BEAGLE library to be present and configured to fully utilize its capabilities.
Bayesian Evolutionary Analysis by Sampling Trees (BEAST) is a cross-platform program for Bayesian inference in evolution and phylogenetics using molecular sequences. 
The current stable release can be obtained from \url{www.beast.bio.ed.ac.uk}.
The Broad-platform Evolutionary Analysis General Likelihood Evaluator (BEAGLE) library is free, open-source software licensed under the GNU LGPL. Both the source code and binary installers are available from \url{www.code.google.com/p/beagle-lib}. 
Scripts and input files required for repeating the simulation study presented in {\bf{Example application}} are hosted at \url{www.github.com/phylogeography/DeepRoot}.

%%%%%%%%%%%%%%5

@@

\begin{figure}[H]
\centering
\includegraphics[scale=0.35]{schema.pdf} 
\caption{
{ \footnotesize 
{\bf Overview of the {\bussname} simulation procedures and GUI screenshot.}
A. Schematic representation of the different ways to employ the {\bussname} simulation software.
Based on an XML input file, simulations can be performed using the core implementation. %in BEAST.
BEAST can parse the specified {\bussname} instructions and generate sequence data as well as analyze the replicate data in a single run.
Using both the GUI or CLI, {\bussname} can run simulations based on an input tree or a list of taxa and their heights.
The software can also write the simulation settings to an XML file that can be then read by BEAST. 
B. The screenshot example shows the set-up of a codon position partitioned simulation in the Partitions panel of the graphical user interface.
The Hasegawa, Kishino and Yano (HKY) model is being set as the substitution model for partition 1, with a $\kappa$ (the transition-transversion bias) parameter value of $4.0$.
}% END: footnotesize 
}
\label{fig:screenshot}
\end{figure}

@@

\textbf{\tiny{Program}} & \textbf{\tiny{Codons\footnote{$\pi$BUSS: GY94, MG94; PhyloSim: GY94 x M0 - M4; Recodon: GY94 x M0, M1, M7, M8; NetRecodon: GY94 x M0, M1, M7, M8; Indelible: GY94 x M0 - M10; Evolver: GY94 x M0, M1, M2, M3, M7, M8; ALF: GY94 x M0, M1, M7 and M8; GenomePop: MG94}}} & \textbf{\tiny{Amino}} & \textbf{\tiny{Indels}} & \textbf{\tiny{Partitions}} & \textbf{\tiny{Molecular}} & \textbf{\tiny{Ancestral}} & \textbf{\tiny{Coalescent}} & \textbf{\tiny{GUI}} & \textbf{\tiny{Multi-}} & \textbf{\tiny{Cross-}}\tabularnewline
 &  & \textbf{\tiny{acids}\footnote{$\pi$BUSS: BLOSUM \cite{Henikoff1992}, CPREV \cite{cprev}, Dayhoff \cite{Dayhoff1978}, FLU \cite{FLU}, JTT \cite{Jones1992}, LG \cite{LG}, MTREV \cite{mtrev}, WAG \cite{WAG}; Seq-Gen: JTT, WAG, PAM \cite{PAM}, BLOSUM, MTREV; indel-Seq-Gen2: PAM, JTT, MTREV, CPREV; PhyloSim: CPREV, JTT, LG , MTART \cite{MTART}, MTMAM \cite{MTMAM}, MTREV24 \cite{MTREV24}, MTZOA \cite{MTZOA}, PAM, WAG; Indelible: Dayhoff, JTT, WAG, VT \cite{VT}, LG, BLOSUM, MTMAM, MTREV, MTART, CPREV, RTREV \cite{rtrev}, HIVb \cite{Nickle2007}, HIVw \cite{Nickle2007}; Evolver: Dayhoff, JTT, WAG, MTMAM, MTREV; ProteinEvolver: BLOSUM, CPREV, Dayhoff, HIVb, HIVw, JTT, Jones \cite{Jones1992}, LG, MTART, MTMAM, MTREV24, RTREV, VT, WAG; ALF: PAM,GTT,LG,WAG; SIMPROT: PAM, JTT, PMB}} &  &  & \textbf{\tiny{clocks}} & \textbf{\tiny{sequences}} &  \textbf{\tiny{models}\footnote{$\pi$BUSS: demography; Recodon: recombination, migration, demography; NetRecodon: recombination, migration, demography; Mesquite: speciation; ProteinEvolver: recombination, migration, demography; SIMCOAL: demography and migration}} &  &  \textbf{\tiny{core}}  &  \textbf{\tiny{platform}\footnote{PhyloSim: R package; Indelible: Executables for Windows and MacOS; ALF: Web interface; GenomePop: Executables for Windows and Linux; SIMCOAL: Executables for Windows; SIMPROT: GUI, Web interface; SIMPROT: Executables for Windows and Linux}}  \tabularnewline

@@

{\tiny{GenomePop \cite{Carvajal2008}}} & {\tiny{X}} &  &  &  &  & {\tiny{X}} & {\tiny{X\footnote{Forward simulation including recombination, demography and migration}}} &  &  & {\tiny{X}}\tabularnewline

@@

\begin{figure}[H]
\centering
\includegraphics[scale=0.6]{deep_root} 
\caption{
{ \footnotesize
{\bf Correspondence between simulated and estimated tMRCAs when purifying selection increases back in time in simulated data sets.} 
}% END: footnotesize
}
\label{fig:deep_root}
\end{figure}

>>>>

\noindent
\textbf{Availability and requirements:} \\
The parallel BEAST/BEAGLE utility for sequence simulation is licensed under the GNU Lesser GPL and its source code is freely available as part of the BEAST Google Code repository: \url{www.code.google.com/p/beast-mcmc/}
Compiled, runnable packages targeting all major platforms along with a tutorial and supplementary data are hosted at:
\url{www.rega.kuleuven.be/cev/ecv/software/pibuss}.
The tutorial is also included as Appendix~\ref{app:pibuss_tuto}.
{\bussname} requires Java runtime environment version 1.5 or later to run its executables and the BEAGLE library to be present and configured to fully utilize its capabilities.
Bayesian Evolutionary Analysis by Sampling Trees (BEAST) is a cross-platform program for Bayesian inference in evolution and phylogenetics using molecular sequences. 
The current stable release can be obtained from \url{www.beast.bio.ed.ac.uk}.
The Broad-platform Evolutionary Analysis General Likelihood Evaluator (BEAGLE) library is free, open-source software licensed under the GNU LGPL. Both the source code and binary installers are available from \url{www.code.google.com/p/beagle-lib}. 
Scripts and input files required for repeating the simulation study presented in {\bf{Example application}} are hosted at \url{www.github.com/phylogeography/DeepRoot}.

@@

\begin{figure}[h!]
\centering
\includegraphics[scale=0.35]{schema.pdf} 
\caption{
{ \footnotesize 
{\bf Overview of the {\bussname} simulation procedures and GUI screenshot.}
A. Schematic representation of the different ways to employ the {\bussname} simulation software.
Based on an XML input file, simulations can be performed using the core implementation. %in BEAST.
BEAST can parse the specified {\bussname} instructions and generate sequence data as well as analyze the replicate data in a single run.
Using both the GUI or CLI, {\bussname} can run simulations based on an input tree or a list of taxa and their heights.
The software can also write the simulation settings to an XML file that can be then read by BEAST. 
B. The screenshot example shows the set-up of a codon position partitioned simulation in the Partitions panel of the graphical user interface.
The Hasegawa, Kishino and Yano (HKY) model is being set as the substitution model for partition 1, with a $\kappa$ (the transition-transversion bias) parameter value of $4.0$.
}% END: footnotesize 
}
\label{fig:screenshot}
\end{figure}

@@

\textbf{\tiny{Program}} & \textbf{\tiny{Codons\footnote{\tiny{$\pi$BUSS: GY94, MG94; PhyloSim: GY94 x M0 - M4; Recodon: GY94 x M0, M1, M7, M8; NetRecodon: GY94 x M0, M1, M7, M8; Indelible: GY94 x M0 - M10; Evolver: GY94 x M0, M1, M2, M3, M7, M8; ALF: GY94 x M0, M1, M7 and M8; GenomePop: MG94}}}} & \textbf{\tiny{Amino}} & \textbf{\tiny{Indels}} & \textbf{\tiny{Partitions}} & \textbf{\tiny{Molecular}} & \textbf{\tiny{Ancestral}} & \textbf{\tiny{Coalescent}} & \textbf{\tiny{GUI}} & \textbf{\tiny{Multi-}} & \textbf{\tiny{Cross-}}\tabularnewline
 &  & \textbf{\tiny{acids}\footnote{\tiny{$\pi$BUSS: BLOSUM \cite{Henikoff1992}, CPREV \cite{cprev}, Dayhoff \cite{Dayhoff1978}, FLU \cite{FLU}, JTT \cite{Jones1992}, LG \cite{LG}, MTREV \cite{mtrev}, WAG \cite{WAG}; Seq-Gen: JTT, WAG, PAM \cite{PAM}, BLOSUM, MTREV; indel-Seq-Gen2: PAM, JTT, MTREV, CPREV; PhyloSim: CPREV, JTT, LG , MTART \cite{MTART}, MTMAM \cite{MTMAM}, MTREV24 \cite{MTREV24}, MTZOA \cite{MTZOA}, PAM, WAG; Indelible: Dayhoff, JTT, WAG, VT \cite{VT}, LG, BLOSUM, MTMAM, MTREV, MTART, CPREV, RTREV \cite{rtrev}, HIVb \cite{Nickle2007}, HIVw \cite{Nickle2007}; Evolver: Dayhoff, JTT, WAG, MTMAM, MTREV; ProteinEvolver: BLOSUM, CPREV, Dayhoff, HIVb, HIVw, JTT, Jones \cite{Jones1992}, LG, MTART, MTMAM, MTREV24, RTREV, VT, WAG; ALF: PAM,GTT,LG,WAG; SIMPROT: PAM, JTT, PMB}}} &  &  & \textbf{\tiny{clocks}} & \textbf{\tiny{sequences}} &  \textbf{\tiny{models}\footnote{\tiny{$\pi$BUSS: demography; Recodon: recombination, migration, demography; NetRecodon: recombination, migration, demography; Mesquite: speciation; ProteinEvolver: recombination, migration, demography; GenomePop: recombination, demography and migration, SIMCOAL: demography and migration}}} &  &  \textbf{\tiny{core}}  &  \textbf{\tiny{platform}
 \footnote{\tiny{PhyloSim: R package; Indelible: Executables for Windows and MacOS; ALF: Web interface; GenomePop: Executables for Windows and Linux; SIMCOAL: Executables for Windows; SIMPROT: GUI, Web interface; SIMPROT: Executables for Windows and Linux}}
 \tabularnewline

@@

{\tiny{GenomePop \cite{Carvajal2008}}} & {\tiny{X}} &  &  &  &  & {\tiny{X}} & {\tiny{X}} &  &  & {\tiny{X}}\tabularnewline

@@

\begin{figure}[h!]
\centering
\includegraphics[scale=0.6]{deep_root} 
\caption{
{ \footnotesize
{\bf Correspondence between simulated and estimated tMRCAs when purifying selection increases back in time in simulated data sets.} 
}% END: footnotesize
}
\label{fig:deep_root}
\end{figure}

### 2015-03-06 chapters/chapter06.tex
changes for printout

\begin{displaymath}
q_{ij}^{MG} = \begin{cases} \theta_{mn,s}\alpha_{s}^{b}\pi_{n}^{p}, & i \to j \text{ is a one-nucleotide synonymous substitution} \\ & \text{from nucleotide $m$ to nucleotide $n$ in codon position $p$,} \\ \theta_{mn,s}\beta_{s}^{b}\pi_{n}^{p}, & i \to j \text{ is a one-nucleotide non-synonymous substitution} \\ & \text{from nucleotide $m$ to nucleotide $n$ in codon position $p$,} \\ 0, & \text{otherwise.} \end{cases}
\end{displaymath}

@@

\begin{figure}[h!]
\centering
\includegraphics[scale=0.2]{continuous.pdf}
\caption{
{ \footnotesize 
{\bf Visualizing continuous phylogeographic reconstructions with uncertainty.} 
A. Example visualization in Google Earth based in estimates obtained from a real data set. B. Conceptual representation of visualizing uncertainty by time slicing a posterior distribution.
} % END: footnotesize
}
\label{fig:cont}
\end{figure}

@@

\begin{figure}
\centering
$$\bullet\; C\;\star C\;\star\; A\;\star\; C\;\star\; T\;\star$$
\caption{
{ \footnotesize 
{\bf DNA sequence CCACT of length $L=5$, with $L+1$ links that separate the characters.} 
By convention to the right of each base there is a \textit{normal link} ($\star$); the leftmost base has an \textit{immortal link} to itself ($\bullet$), ensuring that an insertion can occur at position $l=0$.
} % END: footnotesize
}
\label{fig:indels}
\end{figure}

>>>>

\begin{displaymath}
\footnotesize{
q_{ij}^{MG} = \begin{cases} \theta_{mn,s}\alpha_{s}^{b}\pi_{n}^{p}, & i \to j \text{ is a one-nucleotide synonymous substitution} \\ & \text{from nucleotide $m$ to nucleotide $n$ in codon position $p$,} \\ \theta_{mn,s}\beta_{s}^{b}\pi_{n}^{p}, & i \to j \text{ is a one-nucleotide non-synonymous substitution} \\ & \text{from nucleotide $m$ to nucleotide $n$ in codon position $p$,} \\ 0, & \text{otherwise.} \end{cases}
}
\end{displaymath}

@@

\begin{figure}[h!]
\centering
\includegraphics[scale=0.2]{continuous}
\caption{
{ \footnotesize 
{\bf Visualizing continuous phylogeographic reconstructions with uncertainty.} 
A. Example visualization in Google Earth based in estimates obtained from a real data set. B. Conceptual representation of visualizing uncertainty by time slicing a posterior distribution.
} % END: footnotesize
}
\label{fig:cont}
\end{figure}

@@

\begin{figure}[H]
\centering
$$\bullet\; C\;\star C\;\star\; A\;\star\; C\;\star\; T\;\star$$
\caption{
{ \footnotesize 
{\bf DNA sequence CCACT of length $L=5$, with $L+1$ links that separate the characters.} 
By convention to the right of each base there is a \textit{normal link} ($\star$); the leftmost base has an \textit{immortal link} to itself ($\bullet$), ensuring that an insertion can occur at position $l=0$.
} % END: footnotesize
}
\label{fig:indels}
\end{figure}
